### 2022-03-29 paper/localpsf.tex
reorganizing intro based on Noemi's suggestions
add hmatrix figure. move figures to intro

We present a fast \emph{matrix free} method for approximating operators, $\Aop$, that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained optimization and inverse problems. By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and $\Aop^T$ to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

@@

The output of our method is a hierarchical matrix (H-matrix). H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank. For more details about H-matrices, see Appendix \ref{app:h_matrix}. H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

Since we are building an H-matrix, it is natural to ask why we do not build the H-matrix directly. The reason is that \emph{classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products}. The basic idea behind our method is constructing the H-matrix via impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the function that results from applying $\Aop$ to a delta distribution (point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to evaluate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

@@

@@

The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of points $x_i \in \Omega$ \cite{PRODCONVLIST}. A common choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. However, even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, are required.

@@

In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format. Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to make the following improvements:
\begin{itemize}
	\item PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while the method we propose is based on an assumption that we call ``normalized local mean displacement invariance.'' This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). 
	\item PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. 
\end{itemize}
In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.

Latest developments and improvements in matrix-free H-matrix construction and approximation are based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, which we do not use. These methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more feasible in the future.

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. Here $L^2(\Omega)'$ denotes the topological dual of $L^2(\Omega)$. The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension formally as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{align*}
\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions. If instead $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution, then the Riesz representative of $\rho$ may not exist, so the above derivation does not make sense. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the last line in the derivation is well-defined. Hence, we use the last line of this derivation to \emph{define} the action of $\Aop$ on the distribution $\rho$ as follows:
\begin{equation}
\label{eq:action_on_distribution}
\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
\end{equation}
Here ``$\Aop \genericdistribution^*$'' is an abuse of notation that means the action of $\Aop$ on the distribution $\rho$, per this definition.

@@

%Our method, described in Section \ref{sec:method}, belongs to a class of methods that approximate an operator by interpolating impulse responses of the operator associated with a scattered collection of points. 
Let $\delta_x$ denote the delta distribution\footnote{Recall that the delta distribution $\delta_x:\mathcal{D}(\Omega)\rightarrow \mathbb{R}$ is the linear functional defined by $\delta_x(v) = v(x)$ for all $v\in \mathcal{D}(\Omega)$.} (i.e., point source, impulse) $\delta_x$, centered at the point $x$.
The \emph{impulse response} of $\Aop$ associated with the point $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
that is formed by applying $\Aop$ to $\delta_x$, then taking the Riesz representation of the result. Using \eqref{eq:action_on_distribution} and the definition of the delta distribution, we also see that $\phi_x$ is the function
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Phi(y,x)$ as a function of $y$.

@@

In this paper, we focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Phi$ is sufficiently regular such that $\phi_x$ is a well-defined function for almost all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained (or approximately contained) in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Phi(y,x) \ge 0$$ for all (or nearly all) $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use our method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
Because of these conditions, we are able to estimate the supports of the impulse responses a-priori by applying $\Aop^T$ \footnote{Recall that the transpose of $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ satisfying $\left(\Aop u\right)(v) = \left(\Aop^T v\right)(u)$ for all $u,v\in L^2(\Omega)$.} to a small number of polynomial functions (Section \ref{sec:intromoments}). We use the support estimates to compute scattered impulse responses in batches (Section \ref{sec:get_impulse_response}, also Figure \ref{fig:batches_intro}). This batching allows us to recover many impulse responses using a small number of operator actions $u \mapsto \Aop u$. The support estimates also allow us to interpolate the impulse responses with a new more sophisticated and accurate interpolation scheme (Sections \ref{sec:local_mean_displacement_invariance} and \ref{sec:approximate_kernel_entries}).

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^N_\mathbf{M} \rightarrow \mathbb{R}^N_{\mathbf{M}^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{v}\mapsto \mathbf{A}^T\mathbf{v}$, and computing these matrix-vector products is costly.
%, and each of these matrix-vector products is costly to compute. 
%For example, some computational procedures possess the Galerkin property, in which case $\mathbf{A}_{ij} = \left(\Aop \febasis_j\right)(\febasis_i)$. 
%Our approximation method is agnostic to the computational procedure for performing these matrix-vector products. 
%What is important is that $\mathbf{A}$ is dense, only available via matrix-vector products, and each of these matrix-vector products is costly to compute. 
Of course, matrix entries can be computed via a matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
%\begin{equation*}
%\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i,
%\end{equation*}
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $N$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which may differ from $\Phi$ due to discretization error. Since the functions $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to show that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Let $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.

>>>>

We present a fast \emph{matrix free} method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^d$ is a bounded domain, $d$ is the spatial dimension, and $L^2(\Omega)'$ is the topological dual space of $L^2(\Omega)$. Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{COVARIANCEOPERATORS}, blurring operators in imaging \cite{BLURRING}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{PDEHESSIAN}.

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and $\Aop^T$ to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

@@

The output of our method is a hierarchical matrix (H-matrix). H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank (see Figure \ref{fig:hmatrix_intro}). For more details about H-matrices, see Appendix \ref{app:h_matrix}. H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification. Since we are building an H-matrix, it is natural to ask why we do not build the H-matrix directly. The reason is that \emph{classical H-matrix construction methods \cite{HMATRIXCONSTRUCTION} require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products}.

@@

The basic idea behind our method is impulse response interpolation. The impulse response, $\phi_x$, associated with a point $x$ is the function that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.5]{bct1.eps}
	\end{center}
	\caption{Hierarchical matrix block structure for the Stokes Hessian. Red blocks are dense. All other blocks are low rank.
	}
	\label{fig:hmatrix_intro}
\end{figure}

@@

The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. A common choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. However, computing each impulse response required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required.

@@

In this paper, we build on previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format. Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to make the following improvements:
\begin{itemize}
	\item PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while the method we propose is based on an assumption that we call ``normalized local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). 
	\item PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. 
\end{itemize}
In exchange for these improvements, the proposed approximation cannot be written as a weighted sum of convolution operators, hence we cannot apply this approximation to vectors via the FFT. However, since our approximation is an H-matrix, it can be applied to vectors efficiently via H-matrix methods.

Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LEXINGPEELINGPROCESS}, which we do not use. These methods have, for instance, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{ILONAHMATRIX,NOEMIHMATRIX}. One issue with this approach is that, while these methods are asymptotically scalable in theory, in practice the required number of matrix vector products is large\footnote{DO CALCULATION}. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. 
%We write $L^2(\Omega)'$ to denote the topological dual of $L^2(\Omega)$. 
The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy,
	\end{align}
\end{subequations}
%\begin{align*}
%\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
%&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
%&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
%\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the last line in the derivation is well-defined. Hence, we use the last line of this derivation to \emph{define} the action of $\Aop$ on the distribution $\rho$ as follows:
\begin{equation}
\label{eq:action_on_distribution}
\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
\end{equation}
Since $\genericdistribution^*$ may not exist, here ``$\Aop \genericdistribution^*$'' is simply a notation that means ``the action of $\Aop$ on the distribution $\rho$,'' per \eqref{eq:action_on_distribution}.

@@

%Our method, described in Section \ref{sec:method}, belongs to a class of methods that approximate an operator by interpolating impulse responses of the operator associated with a scattered collection of points. 
Let $\delta_x$ denote the delta distribution\footnote{Recall that the delta distribution $\delta_x:\mathcal{D}(\Omega)\rightarrow \mathbb{R}$ is the linear functional defined by $\delta_x(v) = v(x)$ for all $v\in \mathcal{D}(\Omega)$.} (i.e., point source, impulse) centered at the point $x \in \Omega$.
The \emph{impulse response} of $\Aop$ associated with $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x := \left( \Aop \delta_x^* \right)^*,
\end{equation}
that is formed by applying $\Aop$ to $\delta_x$, then taking the Riesz representation of the resulting linear functional. Using \eqref{eq:action_on_distribution} and the definition of the delta distribution, we see that $\phi_x$ may also be written as the function
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) = \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Phi(y,x)$ as a function of $y$.

@@

In this paper, we focus on approximating operators that satisfy the following conditions: 
\begin{enumerate}
	\item The kernel $\Phi$ is sufficiently regular such that $\phi_x$ is a well-defined function for all $x\in \Omega$.
	\item The supports of the impulse responses $\impulseresponse_x$ are contained in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Phi(y,x) \ge 0$$ for all $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use our method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{enumerate}
Our method may still perform well if these conditions are relaxed slightly. For example, it is fine if the support of the impulse response $\impulseresponse_x$ is not perfectly contained in a localized region, so long as the bulk of the ``mass'' of the impulse response is contained in a localized region. The integral kernel does not need to be non-negative for all pairs of points $(y,x) \in \Omega \times \Omega$, as long as it is non-negative for the vast majority of pairs of points $(y,x)$, and as long as the negative numbers are comparatively small. If these conditions are violated, our method will incur additional error. If these conditions are violated too much, our method may fail. Not all operators of interest (and in particular not all Hessians) satisfy these conditions, but many operators of practical interest do satisfy these conditions, and our method is highly effective for approximating these operators.
%Because of these conditions, we are able to estimate the supports of the impulse responses a-priori by applying $\Aop^T$ \footnote{Recall that the transpose of $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ satisfying $\left(\Aop u\right)(v) = \left(\Aop^T v\right)(u)$ for all $u,v\in L^2(\Omega)$.} to a small number of polynomial functions (Section \ref{sec:intromoments}). We use the support estimates to compute scattered impulse responses in batches (Section \ref{sec:get_impulse_response}, also Figure \ref{fig:batches_intro}). This batching allows us to recover many impulse responses using a small number of operator actions $u \mapsto \Aop u$. The support estimates also allow us to interpolate the impulse responses with a new more sophisticated and accurate interpolation scheme (Sections \ref{sec:local_mean_displacement_invariance} and \ref{sec:approximate_kernel_entries}).

@@

After discretization, the operator $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$ is replaced by an operator $A_h:V_h \rightarrow V_h'$, which becomes an operator
\begin{equation*}
\mathbf{A}:\mathbb{R}^N_\mathbf{M} \rightarrow \mathbb{R}^N_{\mathbf{M}^{-1}}
\end{equation*}
under the isomorphism discussed above. Our method is agnostic to the computational procedure for approximating $\Aop$ with $\mathbf{A}$. What is important is that we do not have direct access to matrix entries $\mathbf{A}_{ij}$. Rather, we have a computational procedure that allows us to compute matrix-vector products $\mathbf{u}\mapsto \mathbf{A}\mathbf{u}$ and $\mathbf{v}\mapsto \mathbf{A}^T\mathbf{v}$, and computing these matrix-vector products is costly.
%, and each of these matrix-vector products is costly to compute. 
%For example, some computational procedures possess the Galerkin property, in which case $\mathbf{A}_{ij} = \left(\Aop \febasis_j\right)(\febasis_i)$. 
%Our approximation method is agnostic to the computational procedure for performing these matrix-vector products. 
%What is important is that $\mathbf{A}$ is dense, only available via matrix-vector products, and each of these matrix-vector products is costly to compute. 
Of course, matrix entries can be computed via matrix-vector products as $\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i$,
%\begin{equation*}
%\mathbf{A}_{ij} = \left(\mathbf{A}\mathbf{e}_j\right)_i,
%\end{equation*}
where $\mathbf{e}_j=(0,\dots,0,1,0,\dots,0)^T$ is the length $N$ unit vector with one in the $j$th coordinate and zeros elsewhere. But computing the matrix-vector product $\mathbf{e}_j \mapsto \mathbf{A}\mathbf{e}_j$ is costly, and therefore wasteful if we do not use other matrix entries in the $j$th column of $\mathbf{A}$. Hence, methods for approximating $\mathbf{A}$ are computationally intractable if they require accessing scattered matrix entries from many different rows and columns of $\mathbf{A}$.

The operator $A_h:V_h \rightarrow V_h'$ can be written in integral kernel form as
\begin{equation}
(A_h u_h)(v_h) := \int_\Omega \int_\Omega v_h(y) \Phi_h(y,x) u_h(x) dx dy
\end{equation}
for some kernel $\Phi_h$ which we do not know, and which may differ from $\Phi$ due to discretization error. Since the functions $V_h$ are continuous at $x$, the delta distribution $\delta_x$ is a continuous linear functional on $V_h$, which has a discrete dual vector $\boldsymbol{\delta}_x \in \mathbb{R}^N_{\mathbf{M}^{-1}}$ with entries $\left(\boldsymbol{\delta}_x\right)_i = \febasis_i(x)$ for $i=1,\dots,N$. Additionally, it is straightforward to verify that the Riesz representation, $\rho_h^* \in V_h$, of a functional $\rho \in V_h'$ has coefficient vector
\begin{equation*}
\boldsymbol{\rho}^* = \mathbf{M}^{-1} \boldsymbol{\rho}.
\end{equation*}
Therefore, the formula for the impulse response from \eqref{eq:impulse_response_delta_action} becomes 
\begin{equation}
\label{eq:discrete_kernel}
\boldsymbol{\impulseresponse}_x = \left(A_h \delta_x^*\right)^* =  \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x,
\end{equation}
and the $(y,x)$ kernel entry of $\Phi_h$ may be written as
\begin{equation}
\label{eq:approx_kernel_entry_81}
	\Phi_h(y,x) = \boldsymbol{\delta}_y^T \boldsymbol{\impulseresponse}_x = \boldsymbol{\delta}_y^T \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \boldsymbol{\delta}_x.
\end{equation}
Now define $\mathbf{\Phi} \in \mathbb{R}^{\fedim \times \fedim}$ to be the following dense matrix of kernel entries evaluated at all pairs of Lagrange nodes:
\begin{equation}
\label{eq:Akerpcmat_entries}
\mathbf{\Phi}_{ij} := \Phi_h(p_i, p_j).
\end{equation}
Because of the Kronecker property of the finite element basis, we have $\boldsymbol{\delta}_{p_i} = \mathbf{e}_i$. Thus from \eqref{eq:approx_kernel_entry_81} we have
$
	\Phi_h(p_i,p_j) = \left(\mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1}\right)_{ij},
$
which implies
\begin{equation}
\label{eq:boldA}
	\mathbf{A} = \bm{M} \mathbf{\Phi} \bm{M}.
\end{equation}
Broadly, our method will construct an H-matrix approximation of $\mathbf{A}$ by forming H-matrix approximations of $\boldsymbol{\Phi}$ and $\mathbf{M}$ (or a lumped mass version of $\mathbf{M}$), then multiplying these matrices with H-matrix methods per \eqref{eq:boldA}. The central challenge is that classical H-matrix construction methods require access to arbitrary matrix entries $\mathbf{\Phi}_{ij}$, but these matrix entries are not easily accessible. The bulk of our method is therefore dedicated to forming approximations of these matrix entries that can be evaluated rapidly.
